% Varieties as Conditioning Structure
% Brett Reynolds
% 2026

\documentclass[11pt,letterpaper]{article}

% House style preamble
\input{.house-style/preamble}

% Document metadata
\title{Varieties as Conditioning Structure:\\
A Game-Theoretic and Bayesian Framework}
\author{Brett Reynolds \orcidlink{0000-0003-0073-7195}\\
  Humber Polytechnic \& University of Toronto\\
  \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}
\date{28 March 2026}

\begin{document}

\maketitle

\begin{abstract}
What distinguishes register from dialect from discourse community? This paper argues that these aren't three kinds of thing but three dimensions of \term{listener expectation}~-- what a listener predicts given their model of the speaker and situation~-- conditioned on different anchors. Register conditions expectation on \textsc{situation} (the here-and-now). Dialect conditions it on \textsc{ascription} (the social categories a speaker is treated as belonging to). Discourse community conditions it on \textsc{identification} (the group whose norms the speaker orients to). The key insight is the distinction between ascription and identification: ascription is what you \emph{are treated as}; identification is who you're \emph{with}. These can align or diverge~-- and the divergence is a major locus of sociolinguistic agency. The framework is projectible: knowing a speaker's S/A/I configuration predicts their distributional behaviour, and observing their behaviour lets you infer S/A/I. I develop this framework by combining \posscite{wiese2023} account of communicative situations as the primary unit of grammatical organization with \posscite{oconnor2019games} game-theoretic explanation of why certain partitions stabilize.
\end{abstract}

% ============================================================================
\section{Introduction}
% ============================================================================

Linguists frequently invoke the concepts of \term{register}, \term{dialect}, and \term{discourse community} to explain variation, but the boundaries between these categories remain blurry. Is a \enquote{legal variety} a register or a discourse community? Is Multicultural Toronto English a dialect or a set of community-anchored registers?

Standard definitions rely on external criteria: dialects are geographic or social, registers are functional or situational. But these criteria often conflict or overlap. A speaker's \enquote{casual register} in one analysis might be another analyst's \enquote{vernacular dialect}. The professional jargon of lawyers might be a register (situational) or a discourse community marker (group-indexed) depending on what question you're asking.

This paper proposes a different approach. Instead of defining variety types by their external correlates, I ask: what does a listener expect from a form~-- and what is that expectation conditioned on?

The answer involves three dimensions. \textsc{Situation}: what's expected in this context. \textsc{Ascription}: what's expected from someone in this social position. \textsc{Identification}: what's expected given the group whose norms the speaker orients to. These map onto register, dialect, and discourse community respectively.

The key insight is the distinction between ascription and identification. Ascription is what you \emph{are treated as}~-- the social categories assigned to you by self and others. Identification is who you're \emph{with}~-- the group whose norms you accept, whose approval you seek. These can align or diverge. A working-class kid in law school has an ascribed position but identifies with a new community. Style-shifting, code-switching, and accommodation all operate in this space between ascription and identification. That space is a major locus of sociolinguistic agency.

Two theoretical sources underpin this framework. From \textcite{wiese2023}: the primary unit of grammatical organization isn't the named language but the \term{communicative situation}~-- the setting of communication as perceived by speakers. From \textcite{oconnor2019games}: the categories that stabilize in a population track \term{payoff space}, not \term{property space}. Varieties emerge as stable strategies in recurring coordination problems.

The S/A/I distinction also provides a conditioning-theoretic gloss on the shifts that variationists themselves often frame in three-waves terms: from macro-categorical modelling of demographic groups, through locally grounded groupings, to third-wave attention to social meaning, style, and stance in interaction \citep[50--55]{drummond_schleef2016_identity}.

Putting these together: register, dialect, and discourse community aren't three kinds of thing. They're three dimensions of listener expectation, distinguished by what the expectation is conditioned on.

This paper does philosophy of sociolinguistics. The contribution is conceptual, not empirical: I'm not presenting new data about any variety but offering a framework for seeing how concepts already in use relate to each other and what that means for practice.

% ============================================================================
\section{Communicative situations as the primary unit}
% ============================================================================

\textcite{wiese2023} argues that grammar doesn't live inside named languages; it lives inside communicative situations. A \term{communicative situation} (com-sit) is the setting of communication as perceived by speakers: a dynamic, socially interpreted bundle of situational characteristics including participants, activity type, medium, goals, and stance.

The key move is that com-sits are primary. Language categories emerge from com-sit distributions; com-sits don't derive from pre-existing language borders.

\subsection{Evidence from free-range settings}

Wiese draws on \enquote{free-range} language settings~-- contexts less constrained by monolingual or standard ideologies~-- to make the underlying organization visible. Urban markets with high linguistic diversity demonstrate translinguistic constructions and emergent local norms. Heritage language settings show com-sit distinctions tied to interlocutors and activities. Multiethnic adolescent peer groups exhibit contact varieties~-- Kiezdeutsch in Berlin, Multicultural Toronto English in Toronto~-- that emerge in tightly defined com-sits and expand as their com-sit base widens.

These settings reveal the conditioning structure precisely because named-language ideologies are weak. The same mechanisms operate in monolingual settings, but the social overlay of language labels obscures them.

\subsection{Acquisition of com-sit conditioning}

Children learn com-sit differentiation early. Wiese describes a child who initially uses a \enquote{Daddy register}~-- German exclusively with father, and only with father~-- then later broadens to a conventional com-sit definition (German in wider contexts). The idiosyncratic stage highlights that com-sit identification is a \emph{learned hypothesis}, not a fixed property of the input.

This acquisition pattern matters for the Bayesian framing developed below: learners track which conditioning variables matter and update their expectations accordingly. The child's initial overly narrow conditioning is later revised as evidence accumulates.

\subsection{Three consequences}

Wiese's framework has three immediate consequences for how we think about variety concepts:

\begin{enumerate}
    \item What we call \textbf{register} is expectation conditioned on \textsc{situation}~-- the conditional reweighting of options induced by the here-and-now. An MTE speaker uses more MTE features with friends than with a teacher; the situation shifts, and the distribution shifts with it.
    \item What we call \textbf{dialect} is expectation conditioned on \textsc{ascription}~-- durable distributional differences tied to social categories speakers belong to. The same speaker's MTE features reflect where they grew up and who they grew up talking to; this baseline persists across situations.
    \item What we call \textbf{discourse community} is expectation conditioned on \textsc{identification}~-- the reference population whose norms the speaker orients to. Whether that speaker treats MTE or Standard Canadian English as the baseline for \enquote{normal} depends on who they're identifying with~-- and that can change.
\end{enumerate}

(I retain \enquote{discourse community} for the third term, though the concept here is broader than its technical use in applied linguistics, where it typically implies shared goals, genres, and gatekeeping practices. The core idea~-- a reference population whose norms one orients to~-- is closer to what sociologists call a \term{reference group} or what Eckert might call a \term{community of practice}. I keep the traditional label for continuity with the variety-types literature.)

But Wiese's framework is primarily descriptive. It tells us that com-sits are the primary unit, but not \emph{why} certain partitions of com-sit space stabilize rather than others. For that, we need game theory.

% ============================================================================
\section{Com-sits as coordination problems}
% ============================================================================

\textcite{oconnor2019games} uses evolutionary game theory to ask: do linguistic terms evolve to track property clusters? Her answer is sobering. Even when property clusters exist, evolved terms may be \emph{conventional}~-- different stable partitions can emerge that don't map cleanly to the natural clusters. And when successful action depends on payoff structure rather than property structure, terms track payoff-relevant distinctions.

The upshot: terms track payoff space, not property space.

\subsection{Communicative situations as payoff structures}

This insight transforms how we think about varieties. Communicative situations are \term{payoff structures}. When speakers choose among variant forms, they're solving \term{coordination problems}: signalling identity, managing politeness, avoiding conflict, claiming authority, marking solidarity.

Consider an MTE speaker navigating a day in Toronto. With neighbourhood friends, using MTE features signals local identity and solidarity~-- the perceived payoff is belonging. With a teacher, a customer at work, or a job interviewer, suppressing those features manages impression~-- the perceived payoff is being taken seriously. Neither choice is more \enquote{authentic}; both are rational responses to different payoff structures.

The partitions that stabilize~-- what we call registers, dialects, and discourse communities~-- are stable strategies in these coordination games. They stabilize not because they mirror some intrinsic clustering of linguistic properties, but because they solve recurring social problems.

This explains why sociolinguistic categories are often messy. Variety boundaries are shaped by payoff relevance (identity, prestige, alignment) rather than by structural clustering alone. Multiple overlapping categories can coexist because different payoff spaces yield different stabilized partitions. Third-wave work in contemporary urban youth settings argues that it's neither possible nor desirable to treat local practice as instantiating one named variety, since features circulate in ways that don't match the patterns outsiders expect \citep[59]{drummond_schleef2016_identity}.

\subsection{Conventionality comes in degrees}

\textcite{oconnor2021-conventionality} develops this further by showing that conventions vary in their degree of arbitrariness. Some conventions are tightly constrained~-- only a few outcomes were ever plausible. Others are highly arbitrary~-- many alternatives could have stabilized.

For linguistic varieties, this means:
\begin{itemize}
    \item Some features are \textbf{low-arbitrariness}: turn-taking repair strategies, articulatory constraints on phonology, information-structural defaults. These are tightly constrained by functional pressures.
    \item Others are \textbf{high-arbitrariness}: specific discourse markers, particular politeness formulae, lexical choices for in-group signalling. These could have been otherwise.
\end{itemize}

In MTE, th-stopping may be relatively low-arbitrariness~-- a common outcome of language contact, emerging independently in many multilingual urban settings. But specific MTE discourse markers or slang items are higher-arbitrariness: they could have been different, and their particularity is what makes them useful for in-group signalling.

The key insight: \enquote{functional doesn't mean inevitable; arbitrary doesn't mean useless.} A feature can be both functionally motivated and historically contingent to different degrees.

\subsection{Categories and social outcomes}

\textcite{oconnor2022contracts} adds a further dimension. When social categories become salient in coordination games, the equilibria that emerge tend to be inequitable~-- discriminatory norms emerge and persist even when most participants would prefer otherwise. Even arbitrary social categories (the \enquote{green bellies} and \enquote{yellow bellies} of her models) produce discriminatory equilibria under standard evolutionary dynamics. History determines \emph{which} categories become salient, but the dynamics themselves generate inequity whenever categories exist. This has significant implications for how variety labels function socially, implications I'll develop in Section~\ref{sec:inequity} after the framework is fully in place.

These game-theoretic results reveal a stabilizer that operates alongside S/A/I but differs from them in kind. S/A/I are \term{exogenous} stabilizers: situation stabilizes because activities recur; ascription because others' perceptions are durable; identification because normative orientation is sticky. Coordination costs are an \term{endogenous} stabilizer: once a community converges on a bundle of conventions, unilateral deviation is costly~-- that's what makes it an equilibrium. \textcite{oconnor2021-conventionality} formalizes this through \term{basins of attraction}: a variety occupies a basin, switching means climbing over the boundary between basins, and a deeper basin absorbs more perturbation. These two kinds of stabilizer interpenetrate~-- ascription locks in partly \emph{because} others coordinate on it; coordination costs operate \emph{through} S/A/I~-- but the distinction is useful: S/A/I are the dimensions variationists already model well; coordination costs are the endogenous dynamics that game theory makes visible. The endogenous stabilizer also explains why varieties are bundles rather than independently varying features: speakers coordinate on the whole package.

% ============================================================================
\section{Varieties as emergent partitions}
% ============================================================================

Combining Wiese and O'Connor yields a unified picture. Varieties are stable partitions of com-sit space, shaped by payoff relevance and sustained by coordination dynamics. But to make this precise, we need to ask: what exactly are we partitioning? What's the fundamental category?

\subsection{Listener expectation as the fundamental category}

I propose that we're dimensioning \term{listener expectation}~-- what a listener predicts given their model of the speaker and context. Listener expectation has multiple facets: a form can be grammatical but situationally unexpected, or phonetically marked but structurally fine. These facets are conditioned differently. Register fit is primarily \textsc{situation}-conditioned: what's expected here and now. Phonetic/accent fit is primarily \textsc{ascription}-conditioned~-- the patterns you bring with you~-- though \textsc{identification} matters when orienting to a new group's norms. Grammaticality is conditioned on both \textsc{ascription} and \textsc{identification}: what counts as grammatical depends on whose data you're drawing from (your community baseline) and whose norms you're orienting to \citep[for extended discussion of these facets]{reynolds2025stack}. For present purposes, the key point is that all facets of listener expectation are conditioned, and the three dimensions below specify \emph{what} they're conditioned on.

Listener expectation is always conditioned on something: a form isn't expected or unexpected simpliciter, but expected \emph{given} some anchor. The traditional variety types differ in what that anchor is.

To make this concrete: listener expectation, as I'll use it, is a listener's predictive response to a form given their inferred conditioning state. A listener hears a form, infers something about the situation, the speaker's social position, and whose norms are in play, then evaluates the form against what they'd predict given those inferences. \enquote{Expected} means: this form is likely given my conditioning; \enquote{unexpected} means: unlikely. The evaluation is probabilistic, gradient, and relative to the listener's model~-- which may differ from the speaker's. This is the sense of listener expectation that the three dimensions below are dimensions \emph{of}.

This yields three dimensions:

\begin{itemize}
    \item \textsc{Situation}: What context are we in? Expectation conditioned on the here-and-now: activity type, medium, formality, audience. This is what \term{register} tracks.
    \item \textsc{Ascription}: What category does the speaker belong to? Expectation conditioned on social position: region, ethnicity, class, generation~-- features that are assigned or inherited, not chosen. This is what \term{dialect} tracks.
    \item \textsc{Identification}: Whose norms is the speaker orienting to? Expectation conditioned on group allegiance: the community whose usage the speaker treats as the baseline for correctness. This is what \term{discourse community} tracks.
\end{itemize}

Of these three, only \textsc{situation} falls squarely within Wiese's com-sit framework. Com-sits are defined by activity type, medium, participants present, goals, stance~-- features of the here-and-now. \textsc{Ascription} is different: it's what the speaker \emph{brings to} the com-sit, their durable social position that conditions behaviour across com-sits. And \textsc{identification} is a learner-side variable~-- whose usage the speaker is tracking as a reference for what's normal. Wiese's framework enables these dimensions but doesn't explicitly theorize them. What we're adding is the claim that listener expectation depends not just on the com-sit but on who the speaker is treated as being and who they're orienting to.

The key distinction is between \textsc{ascription} and \textsc{identification}. Ascription is what you \emph{are treated as}~-- the social categories assigned to you by self and others, based on background, history, or perceived membership. It's a classification process, not a metaphysical essence: ascriptions can be contested, strategically managed, and revised over time, but they have material consequences because others condition their behaviour on them. This keeps \textsc{ascription} usable as an analyst's handle on socially consequential classification without presupposing that members of any descriptive category share a single stable identity with an essential link to language \citep[52]{drummond_schleef2016_identity}. Identification is who you're \emph{with}~-- the group whose norms you accept, whose approval you seek, whose usage you try to match. These can align (you identify with your ascribed community) or diverge (you orient to a community you weren't born into).

This divergence is a major locus of sociolinguistic agency. A working-class kid in law school has an ascribed position (class background, regional origin) but identifies with a new community (legal profession). Their speech reflects both: the ascribed baseline modulated by the identificational target. Similarly, a speaker from a multiethnic Toronto neighbourhood may be ascribed to Multicultural Toronto English but identify differently depending on context~-- embracing it with friends, orienting toward Standard Canadian English with a teacher, or code-switching fluidly across the day. Style-shifting, code-switching, passing, and accommodation all operate in this space between ascription and identification.

\subsection{Stability profiles}

The three dimensions aren't parallel in type~-- they're different \emph{kinds} of conditioning variables, operating at different levels and with different update dynamics:

\begin{itemize}
    \item \textbf{Situation} is token-level: it shifts rapidly~-- within a day, within a conversation, sometimes within an utterance. It's partly determined by external context, but participants also \emph{construct} activity frames, ratify roles, and renegotiate footing~-- situation is an interactional achievement, not just a given.
    \item \textbf{Ascription} is speaker-level: it's durable, tracking features that others (and self) treat as stable properties of the speaker. Ascriptions are often externally assigned and hard to shake, though they can be contested and may shift over biographical time.
    \item \textbf{Identification} is trajectory-level: it's negotiable but sticky. You can change who you orient to, but such changes are effortful and socially consequential. Identification can also be audience-specific~-- you might orient to different norm centres with different interlocutors~-- and it can be multi-layered, with competing identifications in tension.
\end{itemize}

This maps onto the traditional intuitions: registers shift fast, dialects are stable, communities are chosen but hard to switch. The framework makes explicit not just that they differ in stability, but \emph{why}: they operate at different levels and have different update dynamics.

\subsection{Where variables live}

The framework also predicts where sociolinguistic variables should be found. Where S, A, and I converge~-- where all three dimensions predict the same form~-- the feature is invariant. It sits in the stable core of the cluster. A feature becomes variable precisely when the conditioning dimensions pull in different directions: situation predicts one form, ascription another, identification a third. The envelope of variation should map onto the divergence structure of the conditioning space. Variables aren't randomly distributed across the grammar; they cluster at the fringes where different S/A/I configurations make different predictions.

This is a testable prediction. If variables showed no relationship to S/A/I divergence~-- if they were distributed randomly with respect to the conditioning structure~-- the framework would be wrong about where variation comes from. In Appendix~A, I illustrate this with a worked example from \posscite{kiesling1998} study of (ING) in a fraternity, where the same variable shows opposite style effects for speakers with different identifications.

\subsection{Preserving working distinctions}

This analysis doesn't collapse the traditional categories. It clarifies what makes them different. Register, dialect, and discourse community remain useful working concepts~-- they pick out real patterns that matter for sampling, stratification, and analysis. What the framework adds is a principled account of \emph{why} they're different: they condition listener expectation on different things.

The practical distinctions hold. If you're studying regional variation, you're conditioning on ascription. If you're studying style-shifting, you're looking at how situation modulates an ascribed baseline. If you're studying professional socialisation, you're tracking how identification shifts. The framework doesn't dissolve these~-- it tells you what you're doing when you make them.

\subsection{Relationship to three-waves variationism}

The three waves of variationist sociolinguistics can be read as successive discoveries of conditioning dimensions. First-wave work found \textsc{ascription}: macro-sociological categories (class, sex, ethnicity) condition variant rates \citep{drummond_schleef2016_identity}. Second-wave work found \textsc{situation}: locally grounded groupings and social networks condition behaviour in ways that demographic categories miss. Third-wave work found \textsc{identification}: speakers deploy variation agentively for identity work, and social meaning is constructed in interaction through stance and style.

Each wave foregrounded its discovery while backgrounding the others. \textcite{drummond_schleef2016_identity} describe the resulting tension: first- and second-wave work treats identity as stable and imposed; third-wave work treats it as fluid and constructed. Both are real simultaneously, but neither tradition can do formal work with the gap between them. S/A/I holds both at once by making the durable dimension (ascription) and the agentive dimension (identification) structurally different parameters whose \emph{divergence} generates specific predictions~-- not just whose coexistence needs acknowledgment.

What does S/A/I add beyond what indexical fields \citep{Eckert2012}, enregisterment \citep{agha2005}, and indexical orders \citep{silverstein2003} already provide? Not the individual distinctions~-- variationists have been making these under different terminology for decades. The contribution is threefold: (i) the formal separability of the durable and the agentive, with their divergence generating specific distributional predictions~-- indexical fields can describe the meanings available to a variant, but they don't predict which direction a style-shift will go for a speaker with a given identification (Appendix~A shows that S/A/I does); (ii)~the game-theoretic explanation of \emph{why} equilibria are inequitable, not just that they are; and (iii)~the explicit link between indexicality and inverse conditioning, which makes social meaning fall out of the production model rather than being added on. Enregisterment is the process that creates S/A/I configurations; indexical fields are the listener-side inference that S/A/I enables. The framework unifies rather than replaces these concepts.

An objection: if dialect indexes ascription, how do we handle stylisation, crossing, and second-dialect acquisition~-- cases where speakers use features not \enquote{theirs}? The answer is that ascription is an inferred category assignment, not a metaphysical property. When a speaker stylises another variety, listeners may update their ascription of that speaker, or may hear the tokens as marked performance against an unchanged ascription. When someone acquires a second dialect through long-term mobility, their ascription shifts~-- others begin treating them as belonging to a different category. The framework predicts exactly this: ascription is what you're treated as, and what you're treated as can change through sustained behaviour, even if such changes are slow and contested. Crossing and stylisation are cases where the speaker's behaviour and the listener's ascription are strategically misaligned~-- and that misalignment is socially meaningful precisely because ascription normally predicts behaviour.

\subsection{Ballroom culture as a case study}

Consider how variety features actually spread. Lexical items like \mention{shade}, \mention{read}, \mention{tea}, \mention{slay}, and \mention{serving} originated in Black and Latino queer ballroom culture. This community involved both ascription (race, sexuality) and identification (participation in ball culture, house membership)~-- real communities often do. The triad doesn't require that communities be purely one-dimensional; it's about which anchor \emph{dominates} listener expectations for particular forms. For ballroom vocabulary, identification was the dominant anchor: these forms were in-group markers, indexing orientation to ballroom norms.

In the game-theoretic framework: using these forms signalled membership and solved coordination problems (who belongs? who has authority to judge?). The forms stabilized because they served payoff-relevant functions for people who identified with that community.

Then the identificational base expanded. \emph{RuPaul's Drag Race} brought ballroom discourse to wider audiences, and the forms began appearing in new contexts~-- first among people identifying with broader queer culture, then stan Twitter, then mainstream social media. What started as identification markers developed situational variation (ironic vs.\ sincere use, in-group vs.\ appropriative use). Some items are now general English.

This trajectory illustrates the framework. The variety dimensions describe how listener expectation is currently conditioned, and that conditioning can shift. Ballroom vocabulary began as identification-conditioned (you used it to claim membership), developed situational conditioning as it spread (expected in some contexts but not others), and for some items is now approaching ascription-like durability as younger speakers acquire it without the original identificational reference~-- it's just how people their age talk.

% ============================================================================
\section{Bayesian formalisation}
% ============================================================================

The game-theoretic account explains \emph{why} certain partitions stabilize. The Bayesian account explains \emph{how} learners track them. The two are complementary: games explain population-level dynamics; Bayes explains individual learning.

The core idea is simple: learners keep track of which forms are expected in which circumstances, and they update their expectations as they accumulate experience. The formal machinery below makes this precise. But a clarification about what this formalisation is and isn't: the equations identify the conditioning variables and their hierarchical structure. They are a precise statement of the conceptual claim, not a fitted model. The contribution is identifying S/A/I as the dimensions that parameterise listener expectation~-- not delivering a computational implementation. A computational model would require specifying priors from data, fitting parameters, and comparing predictive accuracy against alternatives. That is the natural next step, but it is not what this paper does. The equations earn their keep by making the framework's commitments explicit and falsifiable, not by computing posteriors.

\subsection{Setup}

Consider a choice point~-- a \term{constructional niche}~-- where speakers choose among alternatives that perform roughly the same communicative function. Think of first-person \mention{mans} vs.\ \mention{I}~-- a niche where MTE speakers have an option that Standard Canadian English speakers lack \citep{denis2016mans}. An MTE speaker might say \mention{mans made it} with friends but \mention{I made it} with a teacher. The form varies by situation, but only MTE speakers have both options.

For each utterance, we can note:
\begin{itemize}
    \item What variant the speaker used
    \item The situational context (formal meeting? casual chat?)
    \item The speaker's social position (where they grew up, what groups they belong to)
    \item Whose usage the speaker seems to be orienting to
\end{itemize}

In notation: $Y$ is the observed choice, $S$ captures situation (the here-and-now), $A$ captures ascription (the speaker's social categories), and $I$ captures identification (whose norms the speaker orients to). The learner's task is to figure out: given all this, how likely is each variant?

\subsection{The three dimensions as conditioning variables}

Each dimension plays a distinct role. For readers comfortable with the notation, I'll give the formal statement; the gloss follows.

\textbf{Situation}: If you know the context, you can predict which variants are likely.
\begin{equation}
    Y \mid S = s \sim \text{Categorical}(\theta_s)
\end{equation}
Gloss: \enquote{The distribution over variants depends on the situation.} With a teacher, a speaker might favour \mention{I}, hoping to come across as professional; with friends, \mention{mans} might feel more natural, signalling solidarity.

\textbf{Ascription}: If you know the speaker's social position, you can predict their baseline tendencies.
\begin{equation}
    Y \mid A = a \sim \text{Categorical}(\theta_a)
\end{equation}
Gloss: \enquote{The distribution over variants depends on what categories the speaker belongs to.} A speaker ascribed to Multicultural Toronto English might show higher rates of th-stopping, first-person \mention{mans}, or certain intonation patterns~-- features that track where they grew up and who they grew up talking to, and that persist across situations.

\textbf{Identification}: If you know whose norms the speaker is orienting to, you can predict what they'll produce.
\begin{equation}
    Y \mid I = i \sim \text{Categorical}(\theta_i)
\end{equation}
Gloss: \enquote{The distribution over variants depends on who the speaker is trying to sound like.} An MTE speaker who orients to Standard Canadian English norms~-- perhaps hoping that sounding \enquote{standard} will help professionally~-- might suppress \mention{mans} even with friends. One who identifies strongly with the MTE community might not. The question isn't what situation they're in or where they're from, but whose approval they're hoping for.

In practice, all three matter at once. The full model conditions on situation, ascription, and identification jointly:
\begin{equation}
    Y \mid S{=}s,\, A_{l,s}{=}a,\, I{=}i \;\sim\; \text{Categorical}(\theta_{s,a,i})
\end{equation}
Gloss: \enquote{The distribution over variants depends on all three dimensions jointly.} This is not three independent effects added together; it's a single distribution parameterised by the full S/A/I configuration. Note that ascription is indexed to the listener-speaker pair ($A_{l,s}$), because different listeners may ascribe the same speaker differently~-- ascription is a property of the perception, not the person.

Marginalise over $A$ and $I$ to recover register variation. Marginalise over $S$ and $I$ to recover dialect variation. Marginalise over $S$ and $A$ to recover discourse-community variation. The three \enquote{variety types} are slices of one joint distribution, not three separate models.

\subsection{Hierarchical structure}

These dimensions interact. A speaker has an ascribed position, chooses (or inherits) an identification, and modulates both by situation. The hierarchical model makes this explicit:
\begin{align}
    \theta_{s,a,i} &\sim \text{Dirichlet}(\alpha_{a,i}) \\
    \alpha_{a,i} &\sim \text{Gamma}(\mu_a, \sigma_a)
\end{align}
Speaker-level parameters $\theta$ are drawn from ascription-level distributions (speakers with similar ascriptions share structure), which are themselves drawn from population-level priors. Situation enters as a predictor at the token level, but its effect can vary by identification~-- allowing the model to learn that situational sensitivity itself differs depending on whose norms a speaker orients to. This is a standard Bayesian multilevel model, not a novel architecture~-- the contribution is identifying S/A/I as the conditioning variables that parameterise it.

For quantitative work, this structure matters. Tokens from the same speaker aren't independent; speakers with similar ascriptions share structure; situational effects operate within these nested levels. The dimensionality of $\theta$ depends on the variable and the community: a binary variable in a community with three ascription groups, two identification orientations, and two situational levels yields 12 $\theta$ vectors. The framework doesn't prescribe the granularity~-- it identifies the conditioning dimensions.

A potential worry: identification shapes production, and production signals identification. But this isn't a vicious circle; it's the homeostatic feedback loop the framework predicts. As long as the model is time-indexed~-- the posterior at $t$ becomes the prior at $t{+}1$~-- reciprocal causation is a feature, not a bug. This is a standard property of sequential Bayesian updating in cognitive science and iterative learning models in the language evolution literature.

\subsection{Indexicality as inverse conditioning}

This framework makes indexical meaning fall out automatically. If certain forms are more likely in certain contexts, then hearing a form tells you something about the context. That's just Bayes' theorem running in reverse.

\begin{itemize}
    \item If \mention{mans} is more common in casual, in-group contexts, then hearing \mention{mans} suggests that kind of situation. The form indexes \textsc{situation}.
    \item If \mention{mans} is more common among speakers from MTE backgrounds, hearing it suggests that background. The form indexes \textsc{ascription}.
    \item If using \mention{mans} signals orientation to MTE norms, using it claims alignment with that community. The form indexes \textsc{identification}.
\end{itemize}

Because indexical links accrue in ideologically structured bundles, a single variant typically supports a field of partially compatible inferences (stances, personae, macro-categories) rather than a one-to-one mapping, with context selecting which region of that field is foregrounded \citep[55--57]{drummond_schleef2016_identity}.

Social meaning isn't an arbitrary add-on. It's the informational flip-side of production conditioning. Listeners infer context from forms because speakers condition forms on context.

Two methodological notes. First, there's a connection to accounts of convention as reproduced precedent. Posterior expectations are the proximate cognitive mechanism by which precedents get weight: if a form has been used successfully in contexts like this, the learner's model gives it higher probability in similar contexts. Ascription and identification function as cues for which precedents to treat as the relevant model set. The Bayesian layer is an implementation of convention-reproduction, not a competing ontology.

Second, operationalising these dimensions requires care. Situation is relatively tractable~-- it can be coded from observable features of the interaction. Ascription is harder but doable~-- speakers can be grouped by region, ethnicity, class, or other sociological categories, ideally with attention to how they're actually perceived by others. Identification is the challenge. If identification is inferred from the very linguistic choices being modelled, the analysis is circular. Independent measures are needed: network structure, self-report, stance markers, participation histories~-- something that doesn't just redescribe the outcome. This is a limitation of any approach that takes identification seriously, and it's a constraint on empirical work, not a reason to abandon the concept. Appendix~A demonstrates one viable strategy: \posscite{kiesling1998} discourse-analytic classification of power orientation functions as an independent measure of identification, and it does explanatory work that speaker demographics alone cannot.

A further limitation: the listener's model is not ideologically neutral. Priors on ascription are shaped by the listener's social position and by ideological structures organised around race, class, sex, age, and other axes of social sorting. The framework can represent this (the prior on $A$ is where ideology enters), but it doesn't derive it. The formation of listener expectations is upstream of the model.

% ============================================================================
\section{Normativity and unacceptability}
% ============================================================================

To what extent does a form sound odd, marked, or wrong? In this framework, the answer is: to the extent that it's unlikely given what the listener expects. The judgment \enquote{that's not how we talk} means: \enquote{given who I think you are, what situation I think we're in, and whose norms I think apply~-- I wouldn't expect that form.}

But we need to distinguish several things that \enquote{sounds wrong} might track. \term{Appropriateness}\footnote{I use \enquote{listener expectation} rather than \enquote{appropriateness} as the framework's organising concept, because \enquote{appropriateness} has been argued to embed normative assumptions about whose communicative practices count as situationally correct. The model-internal concept is what the listener's model predicts given inferred S/A/I, not what is normatively correct. In this section, \enquote{appropriateness} appears in its technical sense~-- probabilistic, listener-relative, and explicitly distinguished from prescriptive \enquote{correctness}.} is the genus: the coupling between a form and the values it signals. \term{Grammaticality} is one species~-- the coupling between morphosyntactic form and structural meaning. It has a fact of the matter: either the zipper closes (form and value mesh) or it doesn't. This is why you can have \emph{illusions} of ungrammaticality~-- garden paths that are well-formed but feel wrong, comparative illusions that are ill-formed but feel fine. Without a fact, \enquote{illusion} makes no sense. The feeling of ungrammaticality is a noisy detector tracking the coupling, not an oracle. Other appropriateness channels~-- register fit, accent, lexical precision~-- have their own form--value couplings, their own detectors, their own failure modes. The S/A/I framework developed here is about what all these channels are \emph{indexed to}: grammaticality is variety-indexed (conditioning on whose grammar), register fit is situation-indexed, and so on. \term{Acceptability} is the phenomenology: what the listener actually experiences, informed by all these channels and by processing factors that can fool any of them. And \term{correctness} is the prescriptive overlay: what gatekeepers enforce, what gets codified and moralized~-- often an ideologized version of one variety's appropriateness norms imposed as if universal.

This makes acceptability~-- and especially the appropriateness component~-- relative to conditioning. The same form can be fine under one set of expectations and jarring under another. th-stopping is unremarkable among MTE speakers but may be noticed~-- and judged~-- by listeners conditioning on Standard Canadian English. A double modal is unremarkable in some Southern American varieties and ungrammatical in others. The form didn't change; the listener's conditioning did.

This account of normativity aligns with what \textcite{pullum2019-normativity} calls the \term{constitutive} view of grammatical rules. Pullum argues that grammatical constraints aren't \term{regulative}~-- they don't tell you what you \emph{ought} to do in some moral sense. They're constitutive: they define what counts as participating in a particular practice. Departing from English grammar isn't like breaking a traffic law~-- it's more like making an illegal move in chess. You're not violating a duty; you're just not playing the game.

But unlike chess, there's no rulebook. Grammatical constraints are \term{emergent} (no one stipulated them), \term{variety-indexed} (different communities play by different rules), and \term{tacit} (speakers follow them without being able to state them). The constitutive force is real~-- deviate far enough and you're not playing the game others are playing~-- but there's no FIDE to consult. Which constraints apply is exactly what the S/A/I framework specifies.

This framing accommodates several observations:

\begin{enumerate}
    \item \textbf{Gradient judgments}: probability is continuous, so judgments can be gradient.
    \item \textbf{Context-dependence}: different conditioning yields different expectations.
    \item \textbf{Norm conflicts}: situations can bring speakers into conflicting expectations~-- especially when ascribed identity norms clash with situational demands. An MTE speaker in a job interview faces exactly this: forms natural to their ascription may be marked in that situation.
    \item \textbf{Innovation}: a novel form becomes learnable when it is attributed to a distinct com-sit cluster rather than treated as noise in the baseline.
\end{enumerate}

% ============================================================================
\section{Varieties, categories, and inequity}
\label{sec:inequity}
% ============================================================================

The framework developed in the preceding sections has implications that are uncomfortable but important. Once we understand varieties as coordination equilibria sustained by category salience, we have to face the fact that these equilibria can be~-- and often are~-- inequitable.

\subsection{Standard language ideology as coordination}

Consider a phenomenon sociolinguists know well: standard language ideology. Despite decades of linguists explaining that all varieties are systematic and expressive, the prestige hierarchy persists. An MTE speaker applying for a professional job in Toronto may find their speech~-- th-stopping, distinctive intonation, certain discourse markers~-- working against them. Students are still corrected for \enquote{non-standard} grammar. Why?

The game-theoretic answer: once \enquote{standard} and \enquote{non-standard} become salient categories, they function as coordination signals. Employers coordinate on using \enquote{standard} speech as a hiring filter~-- not because any individual employer decided to discriminate, but because everyone expects everyone else to use this signal, and deviating is costly. The equilibrium is self-sustaining: even employers who know the linguistic facts face pressure to conform to the coordination, because unilateral deviation puts them at a disadvantage.

The minimal conditions for this are simple: (1) individuals recognize social categories, (2) they condition behaviour on category membership, and (3) they adapt to improve their own payoff. Once these conditions are met, coordination can stabilize on unequal outcomes~-- and it can do so even when most participants would prefer a fairer arrangement.

A caveat on the \enquote{payoff} idiom: the game-theoretic framing can make centuries of institutional sedimentation sound like momentary choice situations. It isn't. The categories that structure coordination~-- race, class, region, language~-- have deep histories. The prestige hierarchy that makes \enquote{standard} speech valuable didn't emerge from a coordination game played yesterday; it was built over generations by colonialism, state formation, mass education, and labour-market restructuring. The game-theoretic point is that once these categories are salient, they \emph{sustain themselves} through coordination~-- but it would be a mistake to think the framework explains how they got salient in the first place. History does that.

\subsection{Why this isn't fatalism}

This analysis might sound like naturalizing discrimination~-- reducing injustice to \enquote{just how coordination works}. But equilibria are contingent on the structure of the game, not laws of nature. Change the structure, and you change what's stable.

The framework is diagnostic, not exculpatory. Understanding \emph{why} inequity persists despite good intentions tells us \emph{where to intervene}. And the diagnosis is specific: if linguistic discrimination is a coordination problem, then purely normative interventions~-- telling people to respect all varieties~-- won't be enough. You can't talk your way out of an equilibrium by changing attitudes alone. The coordination structure has to change.

This actually validates what critical sociolinguists have long argued: that individual attitude change is insufficient without structural change. The game-theoretic framing makes this precise. Exhorting fairness doesn't shift equilibria. Anonymising applications, changing institutional policies, removing the signals that coordination locks onto~-- these change the game itself. But removing one signal doesn't guarantee success if other signals remain available; effective intervention requires attending to the full landscape of coordination anchors.

\subsection{The role of category salience}

A key lever is category salience itself. Inequitable equilibria require that categories be recognized and conditioned on. If the category becomes less salient~-- less available as a coordination signal~-- the equilibrium weakens.

This explains something puzzling: why linguistic discrimination sometimes decreases across generations even without explicit intervention. If younger speakers don't recognize a dialect boundary that their grandparents did, the coordination that sustained discrimination against that variety loses its anchor. The category fades, and with it the equilibrium.

It also explains why making categories \emph{more} salient can sometimes backfire. But this claim needs careful bracketing. Salience operates differently in different domains:

\begin{itemize}
    \item \textbf{Salience for mobilisation}: Making a category visible can be necessary for collective action, solidarity, and resistance. You can't fight discrimination against a group if the group isn't recognized.
    \item \textbf{Salience as sorting heuristic}: The same visibility can enable discrimination when categories become coordination signals for gatekeeping~-- in hiring, housing, institutional access.
    \item \textbf{Institutional vs.\ cultural salience}: Categories encoded in forms, checkboxes, and official rubrics may have different effects than categories circulating in everyday talk and awareness.
\end{itemize}

The framework doesn't say salience is bad. It says salience is a double-edged tool: it enables both solidarity and sorting. In some contexts, de-emphasizing category boundaries may reduce discrimination; in others, making categories visible is essential for resistance. The question is always: salience \emph{for what purpose}, \emph{in what domain}, \emph{controlled by whom}?

\subsection{Diffusion, appropriation, and the ecological tradeoff}

The ballroom case from Section~4 illustrates a phenomenon often discussed as cultural appropriation: linguistic forms originating in a marginalized community diffuse to broader populations, losing their community-specific indexicality along the way.

The positive mechanics are clear. When a community's markers spread widely: (1) the forms lose their ability to function as in-group signals, (2) the conditioning structure shifts as forms become associated with new com-sits, and (3) the originating community's conditioning gets overwritten in broader usage. This is diffusion~-- the same process by which any innovation spreads.

But there's a normative tension here, and O'Connor's framework clarifies what it is.

On one hand, category salience is what discrimination coordinates on. If \enquote{ballroom community member} becomes less salient as a category~-- if the forms no longer index that community~-- then discrimination against that community loses one of its coordination anchors. By O'Connor's logic, reduced salience should reduce inequity.

On the other hand, category salience is also what enables community identity, solidarity, in-group coordination, and pride. When markers diffuse and lose their indexical force, the community loses a coordination resource it built.

This is a genuine tradeoff, not a mistake on either side. Communities aren't wrong to value salience; they're also not wrong to recognize what salience enables. The framework doesn't resolve the dilemma~-- it clarifies its structure.

Category diversity also has system-level costs and benefits.

The benefits of diversity are real. Linguistic innovation often emerges in marginal communities~-- ballroom vocabulary is exactly such a case~-- and enriches the broader repertoire. Multiple communities mean multiple perspectives and challenges to assumptions. Diverse systems are more adaptive; monocultures are efficient but fragile.

The costs of diversity are also real. Coordination is harder with more categories. Conflict is more likely. Transaction costs of navigating difference are non-trivial.

This is an ecological framing. Monocultures are efficient in stable environments but vulnerable to disruption. Diverse ecosystems are less efficient but more resilient and generative. A system that dissolved all category distinctions would reduce the coordination anchors for discrimination~-- but would also lose the niches where innovation emerges.

The question, then, isn't whether appropriation is good or bad in the abstract. It's: how much category differentiation should a system sustain, given that differentiation has both costs (coordination friction, conflict, potential discrimination) and benefits (innovation, adaptability, richness)? That's a collective tradeoff, not one to be resolved by minority communities alone.

\subsection{Implications for the conditioning-structure framework}

For the variety concepts in this paper, the inequity analysis means that category labels~-- \enquote{native speaker}, \enquote{standard}, dialect names, register labels like \enquote{professional}~-- aren't neutral descriptors. Once salient, they become coordination loci for unequal social outcomes.

Stigma and prestige are stable equilibria, not noise or mere prejudice. Register policing (\enquote{professional} language as gatekeeping) is an equilibrium outcome~-- which means it won't yield to appeals alone. It requires changing the coordination structure: what signals institutions use, what categories are available for conditioning, what the payoff landscape looks like for deviation.

The conditioning-structure framework provides a precise vocabulary for this. Discrimination operates through conditioning: institutions condition access on variety membership, and that conditioning stabilizes because everyone expects it. To disrupt the equilibrium, you have to disrupt the conditioning~-- either by changing what's conditioned on, or by changing the consequences of the conditioning.

Progressive language policing~-- enforcing inclusive terminology, sanctioning outdated usage~-- operates through the same coordination mechanism as traditional prestige policing. Both are coordination phenomena: certain forms become signals of group membership, deviation is costly, and equilibria form. But the structural position is not symmetric: prestige norms are enforced by institutions that control material outcomes (hiring, education, housing); counter-norms are typically enforced by communities that don't. The asymmetry is structural, not merely ethical. The value of the framework is that it clarifies the mechanism so that normative argument can proceed on solid ground~-- while recognising that identical mechanisms operating in different structural positions produce profoundly different consequences.

One practical implication follows from a simple observation: people use the information they have. When information is scarce, crude signals are valuable; when information is rich, any single signal's marginal value drops. Enrich the information environment and the coordination anchor loses its grip.

But O'Connor's logic implies that coordination will find \emph{something} to anchor on. You can't eliminate it; you can only redirect it. The normative question, then, is: redirect toward what? The framework suggests two criteria. First, toward properties that are \emph{situation-relevant}~-- conditioning on height for basketball or gregariousness for event organizing adds value; conditioning on \mention{mans} in a service encounter doesn't. Second, toward properties that are \emph{not ascribed}~-- conditioning on what people choose is less troubling than conditioning on what they're assigned. In S/A/I terms: shift conditioning away from \textsc{ascription} toward \textsc{situation} (what the situation actually requires) and \textsc{identification} (what people have agency over). The goal is designing environments where coordination tracks the right things. And we can keep teaching Standard varieties to those who want to learn them~-- offering a repertoire expansion is consistent with valuing identification; the problem is conditioning access on already having it.

% ============================================================================
\section{Conclusion}
% ============================================================================

What distinguishes register from dialect from discourse community? Not the external correlates (situation vs.\ geography vs.\ group membership), but what listener expectation is conditioned on: \textsc{situation}, \textsc{ascription}, or \textsc{identification}.

\begin{itemize}
    \item Register is expectation conditioned on the here-and-now~-- what's predicted in this context.
    \item Dialect is expectation conditioned on ascribed categories~-- what's predicted from someone in this social position.
    \item Discourse community is expectation conditioned on identification~-- what's predicted given the group whose norms the speaker orients to.
\end{itemize}

The key insight is the distinction between ascription and identification. Ascription is what you \emph{are treated as}; identification is who you're \emph{with}. These can align or diverge~-- and the divergence is a major locus of sociolinguistic agency.

This framework emerges from combining two insights. From \textcite{wiese2023}: the communicative situation is the primary unit of grammatical organization, and variety concepts describe regularities across com-sits. From \textcite{oconnor2019games,oconnor2021-conventionality,oconnor2022contracts}: com-sits are coordination problems, the partitions that stabilize track payoff relevance, and category salience can stabilize inequity.

The value of this framework is not that it provides a cleaner taxonomy. It's that the taxonomy is \term{projectible}: knowing a speaker's S/A/I configuration predicts their distributional behaviour, and observing their behaviour lets you infer S/A/I. The framework generates specific predictions: (a)~distributional predictions~-- form frequencies given S/A/I; (b)~acquisition predictions~-- learners track S/A/I as conditioning variables and revise with experience; (c)~diffusion predictions~-- forms spread along identification networks before becoming ascription-durable; (d)~inequity predictions~-- discrimination anchors on salient ascription categories and can be disrupted by shifting conditioning toward situation and identification; (e)~variation predictions~-- sociolinguistic variables should cluster at the divergence points of the conditioning space, where S, A, and I make different predictions (Appendix~A).

Are varieties HPC kinds, or merely conventional partitions? The distinction matters. A merely conventional partition~-- driving on the right~-- has no feedback between the convention and its predictive success. Left would work as well as right; nothing about driving rewards the specific partition beyond the fact that others have adopted it. A variety is different. The clustering generates predictions: knowing the S/A/I configuration lets you predict distributional behaviour. And the predictions sustain the clustering: speakers converge because coordination succeeds, and success depends on the specific forms being bundled together. Deviate on one feature and you've signalled you might be playing a different game~-- the prediction fails and the coordination cost rises. That reciprocal feedback~-- clustering sustains mechanisms, mechanisms sustain clustering~-- is what \enquote{homeostatic} means. Varieties are conventional enough to be coordination equilibria (they could have been otherwise), but structured enough to be projectible (the specific partition matters because it supports induction). They sit in exactly the space HPC theory was designed to describe: real without essences, maintained by mechanisms, projectible for purposes.

The traditional variety concepts remain useful working categories~-- they pick out real patterns that matter for analysis. What the framework adds is a principled account of what makes them different and how they relate. Register, dialect, and discourse community aren't three kinds of thing. They're three dimensions of listener expectation, distinguished by what the expectation is conditioned on and how stable that conditioning is.

% ============================================================================
\appendix
\section{Worked example: (ING) in \posscite{kiesling1998} fraternity}
\label{app:kiesling}
% ============================================================================

\subsection{The variable and the published analysis}

\textcite{kiesling1998} recorded 37 hours of interaction in an American college fraternity and analysed the (ING) variable~-- the alternation between velar [ɪŋ] and alveolar [ɪn] in words like \mention{going}/\mention{goin'}. He coded 1,098 tokens across 11 speakers for speaker, activity type (meeting, socializing, interview), following phonological environment, and grammatical category. All speakers were white fraternity brothers from roughly similar class backgrounds.

The overall pattern was conventional: socializing favoured [ɪn] (.72 Varbrul probability), meetings disfavoured it (.30). Most speakers shifted in the expected direction~-- more standard [ɪŋ] in the formal meeting context. But the speaker~$\times$~activity cross-tabulation (Kiesling's Table~3) reveals striking divergence. Mack~-- the clearest standard-pattern case~-- went from 73\% [ɪn] socializing to 13\% in meetings. Speed, by contrast, barely shifted at all: 95\% to 82\%. And Mick \emph{reversed}: 67\% to 71\%, using slightly \emph{more} [ɪn] in the formal setting. Waterson was categorical [ɪn] in meetings, though with only four tokens (two of which were \mention{fuckin'}, which is almost categorically [ɪn] for all speakers). For progressive verb forms specifically (Kiesling's Figures~7--8), the reversal is even clearer: Speed and Mick both \emph{increase} their [ɪn] in meetings, while other speakers decrease it.

\subsection{Mapping onto S/A/I}

Kiesling's factor groups map directly onto S/A/I:

\begin{itemize}
    \item \textsc{Situation} = activity type (meeting vs.\ socializing vs.\ interview). This is the style dimension.
    \item \textsc{Ascription} = speaker demographics (region, class, ethnicity, fraternity membership). On this dimension, all speakers are essentially identical: white fraternity brothers at the same university.
    \item \textsc{Identification} = whose norms the speaker orients to. This is where Kiesling's analysis does its work. He shows through detailed discourse analysis that the reversers orient to \emph{physically powerful, working-class} alignment roles (hard work, confrontation, bodily authority), while the standard-pattern speakers orient to \emph{structurally powerful} roles (institutional hierarchy, formal authority).
\end{itemize}

The original Varbrul analysis codes speaker as a single factor group, conflating ascription and identification. Because all speakers share the same ascription, the speaker effect \emph{is} the identification effect~-- but the analysis can't see this because it doesn't separate A from I.

\subsection{Predictions}

S/A/I generates three predictions for this case:

\begin{enumerate}
    \item \textbf{A/I separation:} Speakers with the same ascription but different identification should show different (ING) rates. This is confirmed: Speed, Mick, and Waterson have the same A as Mack, Hotdog, Ram, etc., but different I, and the distributions diverge dramatically~-- especially in the meeting context.

    \item \textbf{S $\times$ I interaction with reversed sign:} The magnitude \emph{and direction} of situational shifting should depend on identification. For structural-power identifiers, formality increases [ɪŋ] (standard Labovian pattern). For physical-power identifiers, formality increases [ɪn] (reversed). This is confirmed in Kiesling's Figure 8: the slopes run in opposite directions for the two groups.

    \item \textbf{Divergence-point prediction:} (ING) is variable in this community precisely because S, A, and I make divergent predictions. S says: use more [ɪŋ] in meetings. A says: no prediction (everyone has the same A). I says: it depends on whose norms you orient to. The variable sits at a divergence point in the conditioning space, which is why it's a variable rather than an invariant feature.
\end{enumerate}

\subsection{What would falsify the framework}

If independently measured identification added no predictive power over ascription and situation~-- if all fraternity members showed the same style-shift pattern regardless of their normative orientation~-- the A/I distinction would have no empirical content for this variable. Kiesling's data show the opposite: identification is doing the explanatory work that ascription cannot.

A standard mixed-effects model with style (S) and demographic group (A) as predictors would predict the wrong direction for Speed, Mick, and Waterson. Adding identification (I)~-- operationalised through Kiesling's discourse-analytic classification of power orientation~-- fixes the prediction.

A caveat about what this demonstration does and doesn't show. It shows \emph{consistency}: the framework can re-describe published data in a way that reveals structure the original analysis treated as unexplained speaker variation. That's necessary but not sufficient. A stronger test would be \emph{prediction}: identify a constructional niche where S/A/I divergence is high and predict in advance that it should be highly variable~-- then go look. The divergence-point prediction (Section~4.3) makes this possible in principle: find a niche where situation, ascription, and identification make conflicting predictions, and the framework says that's where variation should live. Testing this against corpus data is the natural next empirical step.

% ============================================================================
\newpage
\printbibliography
% ============================================================================

\end{document}
